
---
title: "decorate: <u>D</u>ifferential <u>E</u>pigenetic <u>Cor</u>egul<u>a</u>tion <u>Te</u>st"
author: "Developed by [Gabriel Hoffman](http://gabrielhoffman.github.io/)"
date: "Run on `r Sys.Date()`"
documentclass: article
output: 
  html_document:
  toc: true
  smart: false
vignette: >
  %\VignetteIndexEntry{decorate: Differential Epigenetic Coregulation Test}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\usepackage[utf8]{inputenc}
---

<!--- 
# run analysis
cd /Users/gabrielhoffman/workspace/repos/decorate/vignettes
# cd /hpc/users/hoffmg01/work/pinnacle/vignettes
# cd ~/work//decorate/
rmarkdown::render("decorate_example.Rmd", output_dir='./', intermediates_dir='./'); system("cat decorate_example.html | grep -v 'Found more than one class' | grep -v 'PythonEmbedInR' > decorate_example2.html")
--->

Standard analysis compares the differences in magnitude in epigenetic signal between two subsets of a dataset.  Here we present a test of differential correlation.

\item 1) Compute correlations between local pairs of features

\item 2) Perform hierarchical clustering of features

\item 3) Produce discrete clusters of features

\item 4) Filter clusters based on strength of correlation

\item 5) Statistical test of differential correlation 

\item 6) Data visualization


```{r load.packages, echo=FALSE, message=FALSE, results='hide'}
suppressPackageStartupMessages(library(knitr))
suppressPackageStartupMessages(library(xtable))

options(xtable.type="html")

knitr::opts_chunk$set(
  echo=FALSE,
  warning=FALSE,
  message=TRUE,
  error = FALSE,
  tidy = FALSE,
  cache = TRUE,
  cache.lazy = FALSE,
  dev = c("png"), 
  fig.width=7, fig.height=7)

options(markdown.HTML.stylesheet = 'css/custom.css')
```


```{r run.analysis, echo=TRUE, message=FALSE, results='hide'}
library(decorate)
library(GenomicRanges)

# load data
data('decorateData')

# Evaluate hierarchical clustering
# adjacentCount is the number of adjacent peaks considered in correlation
treeList = runOrderedClusteringGenome( simData, simLocation, adjacentCount=500 )

# Choose cutoffs and return clusters using multiple values for meanClusterSize 
treeListClusters = createClusters( treeList, method = "meanClusterSize", meanClusterSize=c( 2, 10, 30, 40, 50) )

# Evaluate strength of correlation for each cluster
clstScore = scoreClusters(treeList, treeListClusters )

# Filter to retain only strong clusters
clustInclude = retainClusters( clstScore, "LEF", 0.30 )

# get retained clusters
treeListClusters_filter = filterClusters( treeListClusters, clustInclude)

# collapse similar clusters
treeListClusters_collapse = collapseCluster( treeListClusters_filter, simLocation)

# Plot correlations and clusters in region defind by query
query = range(simLocation)  
```

```{r plot2Decorate, fig.width=15}
fig1 = plotDecorate( treeList, treeListClusters, simLocation, query, showTree=FALSE, showGenes=FALSE )
fig2 = plotDecorate( treeList, treeListClusters_filter, simLocation, query, showTree=FALSE, showGenes=FALSE )
library(cowplot)     
plot_grid( fig1, fig2, ncol=2, labels=c('A', 'B') )
```

## Filter and collase
```{r plotCollapse, echo=TRUE}
plotDecorate( treeList, treeListClusters_collapse, simLocation, query, showTree=FALSE, showGenes=FALSE )
```

### Run test of differential correlation
```{r compare, echo=TRUE, message=FALSE, results="hide"}
# Evaluate Differential Correlation between two subsets of data
# increase npermute on real data
sledRes = evalDiffCorr( simData, metadata$Disease, simLocation, treeListClusters, npermute=c(20, 200))

# get summary of results
df = summary( sledRes )

# print results
head(df)
```
```{r print.table}
head(df)
```

### Compare correlation structure in two subsets of the data
```{r plotCompareCorr, fig.width=4}
# extract peak ID's from most significant cluster
peakIDs = getFeaturesInCluster( treeListClusters, df$chrom[1], df$cluster[1], df$id[1])

# plot comparison of correlation matrices for peaks in peakIDs
#  where data is subset by metadata$Disease
main = paste0(df$chrom[1], ': cluster ', df$cluster[1])
plotCompareCorr( simData, peakIDs, metadata$Disease) + ggtitle(main)
```







