
---
title: "decorate: <u>D</u>ifferential <u>E</u>pigenetic <u>Cor</u>ela<u>a</u>tion <u>Te</u>st"
author: "Developed by [Gabriel Hoffman](http://gabrielhoffman.github.io/)"
date: "Run on `r Sys.Date()`"
documentclass: article
output: 
  html_document:
  toc: true
  smart: false
vignette: >
  %\VignetteIndexEntry{decorate: Differential Epigenetic Corelation Test}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\usepackage[utf8]{inputenc}
---

<!--- 
# run analysis
cd /Users/gabrielhoffman/workspace/repos/decorate/vignettes
# cd /hpc/users/hoffmg01/work/pinnacle/vignettes
# cd ~/work//decorate/
rmarkdown::render("decorate_example.Rmd", output_dir='./', intermediates_dir='./'); system("cat decorate_example.html | grep -v 'Found more than one class' | grep -v 'PythonEmbedInR' > decorate_example2.html") 

scp decorate_example2.html minerva:/hpc/users/hoffmg01/www/software/decorate/decorate_example.html 

# move simulation to website
cd /hpc/users/hoffmg01/www/software/decorate
cp /sc/orga/projects/psychencode/gabriel/decorate_analysis/KIRC2.html .
mv KIRC2.html KIRC.html
cp /sc/orga/projects/psychencode/gabriel/decorate_analysis/EpiMap2.html .
mv EpiMap2.html EpiMap.html
cp /sc/orga/projects/psychencode/gabriel/decorate_analysis/methyl_scz_decorate2.html .
mv methyl_scz_decorate2.html methyl_scz_decorate.html



--->

Standard analysis compares the differences in magnitude in epigenetic signal between two subsets of a dataset.  The decorate workflow evaluates the local correlation structure between nearby epigenetic features and identify clusters of features that are differential correlation between two subsets of a dataset.  Epigenetic datasets where decorate can be applied include ChIP-seq, ATAC-seq and DNA methylation.

decorate v`r packageVersion("decorate")`

# Workflow
\item 0) Data normalization and covariate correction

    + We assume this has already been perform using standard methods for your assay
    + decorate should be run on residuals after removing the effects of confounding variables

\item 1) Compute correlations between local pairs of features

    + `runOrderedClusteringGenome()`

\item 2) Perform hierarchical clustering of features

    + built into `runOrderedClusteringGenome()`

\item 3) Produce discrete clusters of features

    + `createClusters():` can take multiple parameter values to return multiple sets of clusters at different resolutions

\item 4) Filter clusters 

    - Based on strength of correlation
      + `scoreClusters():` evaluate the strength of the correlation structure of each cluster
      + `retainClusters():` identify clusters that pass a cutoff for strength of the correlation structure.  By default filters by LEF, the lead eigen-value fraction, which is the fraction of variance explained by the first eigen-value.  Higher LEF values indicate a stronger correlation structure.
      + `filterClusters():` apply filter to clusters based on this cutoff

    - Identify overlapping clusters and drop if redundant
      + `collapseClusters():` Collapse redundant clusters identified from using multiple parameter values in step (3).  Redundancy of two clusters is evaluated using the Jaccard index, the fraction of epigenetic features shared between two clusters.

\item 5) Statistical test of differential correlation 

    + `evalDiffCorr():` implements 16 tests of differential correlation with different properties.  See 


    run permutations tests in parallel on multicore machine using the sparse leading eigen-value driven test (sLED).
    + the amount of regularization is determind with the parameter `sumabs.seq`, where a value of 1 is the least sparse and decreasing values result in more regularization.
    + see sLED [package](https://github.com/lingxuez/sLED) and [manuscript](https://projecteuclid.org/euclid.aoas/1507168849) for details of `adj.beta`, `rho` and `sumabs.seq` 
    + merge all results into one data.frame with `combineResults()`

\item 6) Data visualization
    + `plotDecorate():` plot local correlation structure, clusters of features and genes in the region
    + `plotCompareCorr():` plot correlation structure for two subsets of the data
    + `plotScatterPairs()`: scatter plot for each pair of features for two subsets of the data 


```{r load.packages, echo=FALSE, message=FALSE, results='hide'}
suppressPackageStartupMessages(library(knitr))

options(xtable.type="html")

knitr::opts_chunk$set(
  echo=FALSE,
  warning=FALSE,
  message=TRUE,
  error = FALSE,
  tidy = FALSE,
  cache = TRUE,
  cache.lazy = FALSE,
  dev = c("png"), 
  fig.width=7, fig.height=7)

options(markdown.HTML.stylesheet = 'css/custom.css')
```

# Analysis of simulated data
```{r run.analysis, echo=TRUE, message=FALSE, results='hide'}
library(decorate)
library(GenomicRanges)
library(cowplot)     
library(EnsDb.Hsapiens.v86)

# load data
data('decorateData')

# load gene locations
# this is ENSEMBL v86 from the hg38 assembly
# but other versions and assemblies are available
ensdb = EnsDb.Hsapiens.v86

# Evaluate hierarchical clustering
# adjacentCount is the number of adjacent peaks considered in correlation
treeList = runOrderedClusteringGenome( simData, simLocation, adjacentCount=500 )

# Choose cutoffs and return clusters using multiple values for meanClusterSize 
# Clusters corresponding to each parameter value are returned 
# and then processed downstream
# By using multiple parameter values, epigenetic features are included in clusters 
# at different resolutions
treeListClusters = createClusters( treeList, method = "meanClusterSize", meanClusterSize=c(20, 30, 40, 50) )

# Evaluate strength of correlation for each cluster
clstScore = scoreClusters(treeList, treeListClusters )

# Filter to retain only strong clusters
# If lead eigen value fraction (LEF) > 30% then keep clusters
# LEF is the fraction of variance explained by the first eigen-value
clustInclude = retainClusters( clstScore, "LEF", 0.15 )

# get retained clusters
treeListClusters_filter = filterClusters( treeListClusters, clustInclude)

# collapse redundant clusters
treeListClusters_collapse = collapseClusters( treeListClusters_filter, simLocation, jaccardCutoff=0.9)

# Plot correlations and clusters in region defind by query
# get single entry giving range of the region
query = range(simLocation)  
```

### Summary of clustering
Number of cluster parameters: `r length(treeListClusters)`

+ `meanClusterSize = c(20, 30, 40, 50)`

Number of total clusters: `r countClusters(treeListClusters)`

+ in `treeListClusters`

Number of clusters after filtering: `r countClusters(treeListClusters_filter)`

+ in `treeListClusters_filter`

Number of clusters after collapsing: `r countClusters(treeListClusters_collapse)`

+ in `treeListClusters_collapse`

## Make plots
### Evaluate correlation structure versus distance
```{r plot.corr.decay}
dfDist = evaluateCorrDecay( treeList, simLocation, verbose=FALSE)

plotCorrDecay( dfDist, outlierQuantile=1e-5 )
```


```{r plot2Decorate.text, eval=FALSE, echo=TRUE}
# A) plot all clusters
plotDecorate( ensdb, treeList, treeListClusters, simLocation, query )

# B) plot clusters after filtering
plotDecorate( ensdb, treeList, treeListClusters_filter, simLocation, query)
```

```{r plot2Decorate, fig.width=15, message=FALSE, results="hide", echo=FALSE}
fig1 = plotDecorate( ensdb, treeList, treeListClusters, simLocation, query)
fig2 = plotDecorate( ensdb, treeList, treeListClusters_filter, simLocation, query)
plot_grid( fig1, fig2, ncol=2, labels=c('A: All clusters', 'B: After filtering') )
```

## Plot after filtering and collapsing redundant clusters
```{r plotCollapse, eval=TRUE, message=FALSE, results="hide", echo=TRUE, eval=FALSE}
plotDecorate( ensdb, treeList, treeListClusters_collapse, simLocation, query)
```

```{r plotCollapse.2, message=FALSE, results="hide", echo=FALSE}
fig = plotDecorate( ensdb, treeList, treeListClusters_collapse, simLocation, query)
plot_grid( fig, ncol=1, labels=c('C: After collapsing clusters'), hjust=0 )
```

### Run test of differential correlation using sLED
Evaluate `treeListClusters_collapse`, the filtered collapsed set of clusters
```{r compare, echo=TRUE, message=FALSE, results="hide"}
library(BiocParallel)
param = SnowParam(4, progressbar=TRUE)

# Evaluate Differential Correlation between two subsets of data
resDiffCorr = evalDiffCorr( simData, metadata$Disease, simLocation, 
  treeListClusters_collapse, method='deltaSLE', BPPARAM=param)

# get summary of results
df = summary( resDiffCorr )

# print results
head(df)
```
```{r print.table}
head(df)    
```
The id column gives the clustering parameter from `runOrderedClusteringGenome()`, so cluster 6 with id 10 is different from cluster 6 with id 50.
Note that n.perm can vary across clusters because decorate uses a data-adaptive method to select the number of permutations.  Decorate devotes more permutations to clusters with smaller p-values 

Combine results to merge properties of each cluster into a single data.frame
```{r combine.results, message=FALSE, echo=TRUE}
df_results = combineResults( resDiffCorr, clstScore, treeListClusters, simLocation)
head(df_results)
```

### Summary of cluster properties
```{r cluster properties, message=FALSE}
# Histogram of LEF
ggplot(df_results, aes(LEF, fill=id)) + geom_histogram(alpha=0.7) + theme_bw(17) + xlim(0,1) + theme(aspect.ratio=1, legend.position="bottom", plot.title = element_text(hjust = 0.5)) + scale_fill_discrete(name = "Requested mean cluster size") + xlab("Lead eigenvalue fraction (LEF)") + ggtitle("Summarize LEF")

# Histogram of mean absolute correlation
ggplot(df_results, aes(mean_abs_corr, fill=id)) + geom_histogram(alpha=0.7) + theme_bw(17) + xlim(0,1) + theme(aspect.ratio=1, legend.position="bottom", plot.title = element_text(hjust = 0.5)) + scale_fill_discrete(name = "Requested mean cluster size") + xlab("Mean absolute correlation") + ggtitle("Summarize absolute correlation")

# Boxplot of number of features per cluster
ggplot(df_results, aes(id, N, fill=id)) + geom_boxplot() + theme_bw(17) + theme(aspect.ratio=1, legend.position="bottom", plot.title = element_text(hjust = 0.5)) + scale_fill_discrete(name = "Feature per cluster") + xlab("Requested mean cluster size") + ylab("Number of features") + ggtitle("Summarize feature per cluster") + coord_flip()
```

### Compare correlation structure in two subsets of the data
```{r plotCompareCorr, fig.width=4, echo=TRUE}
# extract feature identifiers from most significant cluster
peakIDs = getFeaturesInCluster( treeListClusters_collapse, df_results$chrom[1], df_results$cluster[1], df_results$id[1])

query = range(simLocation[names(simLocation) %in% peakIDs])

locText = with( data.frame(query), paste0(seqnames, ':', format(start, big.mark=','), '-', format(end, big.mark=',')))

# plot comparison of correlation matrices for peaks in peakIDs
#  where data is subset by metadata$Disease
main = paste0(df$chrom[1], ': cluster ', df$cluster[1], '\n', locText)   
plotCompareCorr( simData, peakIDs, metadata$Disease) + ggtitle(main)
```

### Compare correlation structure along genome
First, plot correlation structure for controls (metadata\$Disease==0).  Then, plot correlation structure for cases (metadata\$Disease==1).

The cluster located at `r locText` has a p-value of `r format(df_results$pValue[1], digits=4)` and a test statistic of `r format(df_results$stat[1], digits=3)`.  The negative sign indicates a loss of correlation in the test set (i.e. Disease==1) compared to the baseline correlation in controls (i.e. Dissease==0)
```{r decoate.sled, eval=FALSE, echo=TRUE, message=FALSE}
# get location of peaks in this cluster  
query = range(simLocation[names(simLocation) %in% peakIDs])

# expand window to include adjacent clusters
window = 2000 
start(query) = start(query) - window
end(query) = end(query) + window

# correlations for Disease == 0
plotDecorate( ensdb, treeList, treeListClusters_collapse, simLocation, query, data=simData[,metadata$Disease==0])

# correlations for Disease == 1
plotDecorate( ensdb, treeList, treeListClusters_collapse, simLocation, query, data=simData[,metadata$Disease==1])
```




```{r plot2Decorate.query, fig.width=15, message=FALSE, results="hide", echo=FALSE}
# get location of peaks in this cluster  
query = range(simLocation[names(simLocation) %in% peakIDs])

# expand window to include adjacent clusters
window = 2000 
start(query) = start(query) - window
end(query) = end(query) + window

fig1 = plotDecorate( ensdb, treeList, treeListClusters_collapse, simLocation, query, data=simData[,metadata$Disease==0])

fig2 = plotDecorate( ensdb, treeList, treeListClusters_collapse, simLocation, query, data=simData[,metadata$Disease==1])
plot_grid( fig1, fig2, ncol=2, labels=c('A: Contols (i.e. Disease==0)', 'B: Cases (i.e. Disease==1)') )
```

### Pairwise scatter plots
```{r scatter, echo=TRUE, message=FALSE, fig.width=15, fig.height=15}
plotScatterPairs( simData, peakIDs, metadata$Disease) + ggtitle(main)
```
